---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
# Load the GEOquery package
#BiocManager::install("GEOquery")
library(GEOquery)

# Download the data
gse <- getGEO("GSE6014", GSEMatrix =TRUE, destdir = getwd())
```


```{r}
# Extract the expression data
# gse <- gse[[1]]

# Extract phenotype data
phenoData <- pData(phenoData(gse[[1]]))

# Subset phenotype data to only M059K rows
phenoData2 <- phenoData[-c(1,4),]
# Subest phenotype data to only cell_line and treatment columns
phenoData2 <- phenoData2[,c(1,2)]

# Clean up and rename values in the title column
cleanednames <- c('untreated', 'treated')
phenoData2$title <- cleanednames

# Rename title column to treatment
colnames(phenoData2) <- c('treatment','geo_accession')

#Phenotype data2 convert values to factor
phenoData2[sapply(phenoData2, is.character)] <- lapply(phenoData2[sapply(phenoData2,is.character)], as.factor)
```

```{r}
# Extract expression data
expr <- exprs(gse[[1]])

# Subset expression data to only M059K rows
expr <- expr[,-c(1,4)]
                         
# Transpose the expression data
#expr <- t(expr)

# Create a data frame with the expression data
#exprsData2 <- as.data.frame(exprsDataT)

# Add the phenotype data to the expression data
#exprsData2$cell_line <- phenoData2$cell_line
```

```{r}
# Get the annotation GPL id (see Annotation: GPL10558)
gpl <- getGEO('GPL96', destdir=".")

# Inspect the table of the gpl annotation object
View(Table(gpl))
```
```{r}
# Extract probe IDs and ENTREZ_GENE_IDs
probeIDs <- Table(gpl)$ID

entrezGeneIDs <- Table(gpl)$ENTREZ_GENE_ID

# Initialize an empty vector to store gene IDs as strings
geneNames <- rep(NA, length(probeIDs))

# Loop through probe IDs and split ENTREZ_GENE_IDs by the delimiter, then concatenate
for(i in 1:length(probeIDs)){
  geneNames[i] <- str_split(entrezGeneIDs[i], pattern = " /// ")[[1]][1]
}

# Assign probe IDs as names to the geneNames vector
names(geneNames) <- probeIDs

# Display the head of the named vector
head(geneNames)

```

Now we can add gene names to our expression file!

```{r}
rownames(expr)<-geneNames[rownames(expr)]
View(expr)
```

Just like the phenotype files, there is no standard format for our expression array file. Typically in gene expression data we need to:
1) Background correct 
2) Normalize 
3) $Log_2$ transform 
4) (optional) mean or median center 
all our values in order to do any further analysis. We can assess how many of these steps have already been completed by checking a quick boxplot of the data:

```{r}
expr2 <- data.frame(t(expr))
boxplot(expr2[,1:10])
```
Determine how many genes had zero counts and remove these genes (rows) from the data set.

```{r}
MinVals <- apply(expr2, 1, min)
sum(MinVals == 0) 
```

```{r}
ExpLog2 <- log2(expr2)
hist(rowMeans(ExpLog2), xlab = "Counts (log2)", main = "Expressed Genes")

SampleMedians <- apply(expr2, 1, median) # Find the median value of each column - 2 means column
GrandMedian <- mean(SampleMedians) # Take the average of those
CorrectionFactors <- GrandMedian - SampleMedians # Calculate correction factor to apply to data
CorrectionFactors
```

```{r}
ExpNorm <- ExpLog2

for(col in colnames(ExpNorm)){
  ExpNorm[, col] <- ExpLog2[, col] + CorrectionFactors[col]
}
```
Do a simple normalization of your data and visualize your normalized data with a boxplot of log2 raw counts and a cluster dendrogram.

```{r}
boxplot(ExpNorm[,1:20], main = "ExpLog2 Data")
```